---
title: "depmap vignette"
author: "Theo Killian"
date: "February 19, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{depmap vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## depmap

This is a brief vignette displaying the data and functions found in the [depmap](https://depmap.org/portal/download/) package.  

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
```

## Preliminaries

To install R/depmap, you need to first install the package. Type (within R) 
install.packages("qtl") (This needs to be done just once.)

```{r}
#install.packages("BiocManager")
#BiocManager::install("tfkillian/depmap")
```

You then load the R/depmap package using the library function:

```{r}
#library("depmap")
```

This needs to be done every time you start R. (There is a way to have the 
package loaded automatically every time, but we won’t discuss that here.)

To get help on the functions and data sets in R (and in R/depmap), use help() 
or ?. For example, to view the help file for the remove.rows.all.nas function, 
type one of the following:

```{r}
help(remove.rows.all.nas)
?remove.rows.all.nas
```

## Data Import

```{r}
#data("gene_effect_corrected")
#"gene_effect_corrected"

```

## Including Plots

You can also embed plots, for example:

```{r}
#library(ggplot2)
#y <-depmap_19Q1_cell_lines$Primary.Disease
#pie <- coord_polar("y", start=0)
#pie
```