---
title: "depmap vignette"
author: "Theo Killian"
date: "February 19, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{depmap vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, include = FALSE, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(collapse = T, comment = "#>")
```

# [depmap](https://github.com/tfkillian/depmap/)

This is a brief demo vignette displaying the data and functions found in 
the [depmap](https://github.com/tfkillian/depmap/) package.  

```{r, echo=FALSE, warning=FALSE}
# This code chunk simply makes sure that all the libraries used here are installed, it will not be shown in the report (notice echo = FALSE).
packages <- c("RCurl", "knitr", "tidyr", "dplyr", "ggplot2", "plotly", "viridis")
if ( length(missing_pkgs <- setdiff(packages, rownames(installed.packages()))) > 0) {
  message("Installing missing package(s): ", paste(missing_pkgs, collapse = ", "))
  install.packages(missing_pkgs)
}
```

## Preliminaries

To install [depmap](https://github.com/tfkillian/depmap/), the [BiocManager](https://cran.r-project.org/web/packages/BiocManager/index.html) 
Bioconductor Project Package Manager is required. If [BiocManager](https://cran.r-project.org/web/packages/BiocManager/index.html) 
is not already installed, it will need to be done so beforehand. Type (within R)
install.packages("BiocManager") (This needs to be done just once.)

```{r, eval=FALSE, echo=T}
install.packages("BiocManager")
BiocManager::install("tfkillian/depmap")
```

The [depmap](https://github.com/tfkillian/depmap/) package can be loaded using 
the library function:

```{r, eval=FALSE, echo=T}
library("depmap")
```

## Data Import

The [depmap](https://github.com/tfkillian/depmap/) package currently contains 
seven datasets (shown below) in *.rda* format. These datasets can be loaded by 
using *data()* after the [depmap](https://github.com/tfkillian/depmap/) package 
has been loaded in RStudio using *library("depmap")*. 

Brief views of [depmap](https://github.com/tfkillian/depmap/) column subsets are
shown. Some of these datasets possess thousands of columns, which may be gene
names or in other cases cell lineages. For convenience, the list of column names 
of these large data sets is restricted to the first 10. Note: "ACH" in the first 
column in some datasets refers to the DepMap ID#. 

### "CCLE_depMap_19Q1_TPM"
CCLE RNAseq gene expression data (log2(TPM+1)). The portal shows expression data 
for only protein coding genes; this download is the full dataset. Contains: 
Genes 57820, Cell Lines 1165, Primary Diseases 33, Lineages 32. First column is 
the DepMap ID# of the cell lineage. The remaining columns consist of the 
numerical dependency score values of are select genes. 

```{r, eval=FALSE, echo=T}
data("CCLE_depMap_19Q1_TPM", package = "depmap")
CCLE_depMap_19Q1_TPM
print("CCLE_depMap_19Q1_TPM data subset")
View(CCLE_depMap_19Q1_TPM[c(1:10)])
```

### "CCLE_RPPA_20180123"
CCLE Reverse Phase Protein Array (RPPA) data. Contains: Genes 214, Cell Lines 
899, Primary Diseases 28, Lineages 28. First column is the common name of the 
cell lineage. The remaining columns consist of the numerical dependency score 
values of are select genes. 

```{r, eval=FALSE, echo=T}
data("CCLE_RPPA_20180123", package = "depmap")
CCLE_RPPA_20180123
print("CCLE_RPPA_20180123 data subset")
View(CCLE_RPPA_20180123[c(1:10)])
```

### "D2_combined_genetic_dependency_scores" 
Genetic Dependency Combined RNAi data (Broad, Novartis, Marcotte). Contains: 
Genes 17309 Cell Lines 712 Primary Diseases 30 Lineages 31. First column is 
select genes knocked out by RNAi, while the remaning columns are the dependency 
scores of the knockout on specific cancer cell lineages. 

```{r, eval=FALSE, echo=T}
data("D2_combined_genetic_dependency_scores", package = "depmap")
D2_combined_genetic_dependency_scores
print("D2_combined_genetic_dependency_scores data subset")
names(D2_combined_genetic_dependency_scores[c(1:10)])
View(D2_combined_genetic_dependency_scores[c(1:10)])
```

### "depmap_19Q1_cell_lines" 
Metadata about cell lines in the 19Q1 release, including mapping between DepMap 
ID and CCLE names. Contains: Genes 0, Cell Lines 1677, Primary Diseases 38, 
Lineages 33. Columns are "DepMap_ID", "CCLE_Name", "Aliases", "COSMIC_ID", 
"Sanger.ID", "Primary.Disease", "Subtype.Disease", "Gender" and "Source"  

```{r, eval=FALSE, echo=T}
data("depmap_19Q1_cell_lines", package = "depmap")
depmap_19Q1_cell_lines
print("depmap_19Q1_cell_lines data subset")
names(depmap_19Q1_cell_lines)
View(depmap_19Q1_cell_lines)
```

### "depmap_19Q1_mutation_calls" 
Merged mutation calls (coding region, germline filtered). Contains: Genes 18755, 
Cell Lines 1601, Primary Diseases 37, Lineages 33. Columns are: "X1", 
"Hugo_Symbol", "Entrez_Gene_Id", "NCBI_Build", "Chromosome", "Start_position",        
"End_position", "Strand", "Variant_Classification", "Variant_Type", 
"Reference_Allele", "Tumor_Seq_Allele1", "dbSNP_RS", "dbSNP_Val_Status", 
"Genome_Change", "Annotation_Transcript", "Tumor_Sample_Barcode", "cDNA_Change",
"Codon_Change", "Protein_Change", "isDeleterious", "isTCGAhotspot", "TCGAhsCnt",
"isCOSMIChotspot", "COSMIChsCnt", "ExAC_AF", "VA_WES_AC", "CGA_WES_AC", 
"SangerWES_AC", "SangerRecalibWES_AC", "RNAseq_AC", "HC_AC", "RD_AC", "WGS_AC", 
"Variant_annotation", "DepMap_ID"  

```{r, eval=FALSE, echo=T}
data("depmap_19Q1_mutation_calls", package = "depmap")
depmap_19Q1_mutation_calls
print("depmap_19Q1_mutation_calls data subset")
names(depmap_19Q1_mutation_calls)
View(depmap_19Q1_mutation_calls)
```

### "gene_effect_corrected"
Batch corrected CERES inferred gene effect matrix derived from CRISPR knock-out
data. Contains: Genes 17634, Cell Lines 558, Primary Diseases 26, Lineages 28. 
First column is the DepMap ID# of the cell lineage. The remaining columns 
consist of the numerical dependency score values of are select genes. 

```{r, eval=FALSE, echo=T}
data("gene_effect_corrected", package = "depmap")
gene_effect_corrected
print("gene_effect_corrected data subset")
names(gene_effect_corrected[c(1:10)])
View(gene_effect_corrected[c(1:10)])
```

### "public_19Q1_gene_cn" 
DepMap WES copy number data. Contains: Genes 23299, Cell Lines 1604, Primary 
Diseases 38, Lineages 33. First column is the DepMap ID# of the cell lineage. 
Columns with numerical dependency score values of are select genes. 

```{r, eval=FALSE, echo=T}
data("public_19Q1_gene_cn", package = "depmap")
public_19Q1_gene_cn
print("public_19Q1_gene_cn data subset")
names(public_19Q1_gene_cn[c(1:10)])
View(public_19Q1_gene_cn[c(1:10)])
```

## Downloading [Broad Institute](https://depmap.org/portal/download/) data directly

If desired, the depmap datasets can also be downloaded in *.csv* format directly 
from the [Broad Institute](https://depmap.org/portal/download/) website. Be 
aware that these files are 1.5GB in total and take some time to download.

```{r, eval=FALSE, echo=T}
#install.packages("RCurl")
library("RCurl", warn.conflicts = FALSE)

# r read csv from url
# allows you to directly download csv file from website
# download depmap_19Q1_mutation_calls.csv
depmap_19Q1_mutation_calls <- read.csv("https://depmap.org/portal/download/api/download/external?file_name=ccle%2Fdepmap-mutation-calls-9a1a.7%2Fdepmap_19Q1_mutation_calls.csv")
depmap_19Q1_mutation_calls

# download public_19Q1_gene_cn.csv
public_19Q1_gene_cn  <- read.csv("https://depmap.org/portal/download/api/download/external?file_name=ccle%2Fdepmap-wes-cn-data-97cc.14%2Fpublic_19Q1_gene_cn.csv")
public_19Q1_gene_cn

# download DepMap-2019q1-celllines.csv
depmap_19Q1_cell_lines  <- read.csv("https://depmap.org/portal/download/api/download/external?file_name=processed_portal_downloads%2Fdepmap-public-cell-line-metadata-183e.3%2FDepMap-2019q1-celllines.csv")
depmap_19Q1_cell_lines

# download gene_effect_corrected.csv
gene_effect_corrected <- read.csv("https://ndownloader.figshare.com/files/14221385")
gene_effect_corrected

# download CCLE_depMap_19Q1_TPM.csv
CCLE_depMap_19Q1_TPM  <- read.csv("https://depmap.org/portal/download/api/download/external?file_name=ccle%2Fdepmap-rnaseq-expression-data-ccd0.12%2FCCLE_depMap_19Q1_TPM.csv")
CCLE_depMap_19Q1_TPM

# download CCLE_RPPA_20180123.csv
CCLE_RPPA_20180123 <- read.csv("https://depmap.org/portal/download/api/download/external?file_name=ccle%2FCCLE_RPPA_20180123.csv")
CCLE_RPPA_20180123

# download D2_combined_genetic_dependency_scores.csv
D2_combined_genetic_dependency_scores <- read.csv("https://ndownloader.figshare.com/files/13515395")
D2_combined_genetic_dependency_scores
```

## Visualisations

Here are a few plots to help illustrate features of the [depmap](https://github.com/tfkillian/depmap/) data 

```{r, echo=FALSE, fig.width=10}
library(ggplot2, warn.conflicts = FALSE) 
library(viridis, warn.conflicts = FALSE)

cancer_types <-unique(depmap_19Q1_cell_lines$Primary.Disease)
num_cancer <-(as.integer(cancer_types))

pie(x=num_cancer, label = "", init.angle = 90, col = plasma(length(cancer_types)))
legend("left", legend=cancer_types, fill=plasma(length(cancer_types)))


```


In order to visualize the full potential of the depmap data, the function shown
below fetches the dependency scores corresponding to a specific gene and cell 
lineage. For example: BRCA1 and `184A1_BREAST` breast cancer lineage.

PSMC2 (Rpt1) was the highest-ranked CYCLOPS candidate in
our original analysis and was also significant in the validation
data set. PSMC2 is part of the 19S regulatory complex of the
26S proteasome, which is responsible for catalyzing the unfold-
ing and translocation of substrates into the 20S proteasome

```{r, eval=FALSE, echo=T}

```

Compare RNAi and CRISPR datasets

Suggestions for the depmap vignette: (1) replicate one or two figures as produced by the online DepMap Data Explorer portal; (2) compute and visualise initially for a single gene, then for all genes, the correlation between shRNAi- and CASPR-based dependency scores; (3) explore the dependency scores for genes with high and low expression (detailed to be discussed with Axelle).

```{r, eval=FALSE, echo=T}

library("ggplot2", warn.conflicts = FALSE)

D2_combined_genetic_dependency_scores$`184A1_BREAST`

h <-gene_effect_corrected[order(gene_effect_corrected$`A1BG (1)`, decreasing=T)[1:5],]
values <- rbind(h$X1, h$`A1BG (1)`)
values
#0.00034

```

Axelle asks: What is the baseline of the expression data?

```{r, eval=FALSE, echo=T}


```

## TCGA Cancer genome atlas

https://cancergenome.nih.gov/researchhighlights/tcgainaction


