---
title: "depmap vignette"
author: "Theo Killian"
date: "February 19, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{depmap vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, include = FALSE, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(collapse = T, comment = "#>")
```

# [depmap](https://github.com/tfkillian/depmap/)

This is a brief demo vignette displaying the data and functions found in 
the [depmap](https://github.com/tfkillian/depmap/) package.  

## Introduction {#sec:intro}

The [depmap](https://github.com/tfkillian/depmap/) package aims to provide a 
reproducible research framework to cancer dependency data described by 
[Tsherniak, Aviad, et al. "Defining a cancer dependency map." Cell 170.3 (2017): 564-576.](https://www.ncbi.nlm.nih.gov/pubmed/28753430). The [depmap](https://github.com/tfkillian/depmap/) 
package should allow a researcher to easily mine dependency data taken from 
this cancer genomic study, explore the data and its statistical properties 
and visually display them.

```{r, echo=FALSE, warning=FALSE}
# This code chunk simply makes sure that all the libraries used here are installed, it will not be shown in the report (notice echo = FALSE).
packages <- c("RCurl", "knitr", "tidyr", "dplyr", "ggplot2", "plotly", "viridis", "tidyverse")
if ( length(missing_pkgs <- setdiff(packages, rownames(installed.packages()))) > 0) {
  message("Installing missing package(s): ", paste(missing_pkgs, collapse = ", "))
  install.packages(missing_pkgs)
}
```

## Preliminaries

To install [depmap](https://github.com/tfkillian/depmap/), the [BiocManager](https://cran.r-project.org/web/packages/BiocManager/index.html) 
Bioconductor Project Package Manager is required. If [BiocManager](https://cran.r-project.org/web/packages/BiocManager/index.html) 
is not already installed, it will need to be done so beforehand. Type (within R)
install.packages("BiocManager") (This needs to be done just once.)

```{r, eval=FALSE, echo=T}
install.packages("BiocManager")
BiocManager::install("tfkillian/depmap")
```

## Data Import {#sec:data}

The [depmap](https://github.com/tfkillian/depmap/) package currently contains 
seven datasets (shown below) in *.rda* format. These datasets can be loaded by 
using *data()* after the [depmap](https://github.com/tfkillian/depmap/) package 
has been loaded in RStudio using *library("depmap")*. 

```{r, eval=FALSE, echo=T}
library("depmap")
#load(package_version())
```

Brief views of [depmap](https://github.com/tfkillian/depmap/) column subsets are
shown. Some of these datasets possess thousands of columns, which may be gene
names or in other cases cell lineages. For convenience, the list of column names 
of these large data sets is restricted to the first 10. Note: "ACH" in the first 
column in some datasets refers to the DepMap ID#. 

### "CCLE_depMap_19Q1_TPM"
CCLE RNAseq gene expression data (log2(TPM+1)). The portal shows expression data 
for only protein coding genes; this download is the full dataset. Contains: 
Genes 57820, Cell Lines 1165, Primary Diseases 33, Lineages 32. First column is 
the DepMap ID# of the cell lineage. The remaining columns consist of the 
numerical dependency score values of are select genes. 

```{r, eval=FALSE, echo=T}
data("CCLE_depMap_19Q1_TPM", package = "depmap")
CCLE_depMap_19Q1_TPM
print("CCLE_depMap_19Q1_TPM data subset column names 1-10")
View(CCLE_depMap_19Q1_TPM[c(1:10)])
```

### "CCLE_RPPA_20180123"
CCLE Reverse Phase Protein Array (RPPA) data. Contains: Genes 214, Cell Lines 
899, Primary Diseases 28, Lineages 28. First column is the common name of the 
cell lineage. The remaining columns consist of the numerical dependency score 
values of are select genes. 

```{r, eval=FALSE, echo=T}
data("CCLE_RPPA_20180123", package = "depmap")
CCLE_RPPA_20180123
print("CCLE_RPPA_20180123 data subset column names 1-10")
View(CCLE_RPPA_20180123[c(1:10)])
```

### "D2_combined_genetic_dependency_scores" 
Genetic Dependency Combined RNAi data (Broad, Novartis, Marcotte). Contains: 
Genes 17309 Cell Lines 712 Primary Diseases 30 Lineages 31. First column is 
select genes knocked out by RNAi, while the remaning columns are the dependency 
scores of the knockout on specific cancer cell lineages. 

```{r, eval=FALSE, echo=T}
data("D2_combined_genetic_dependency_scores", package = "depmap")
D2_combined_genetic_dependency_scores
print("D2_combined_genetic_dependency_scores data subset column names 1-10")
names(D2_combined_genetic_dependency_scores[c(1:10)])
View(D2_combined_genetic_dependency_scores[c(1:10)])
```

### "depmap_19Q1_cell_lines" 
Metadata about cell lines in the 19Q1 release, including mapping between DepMap 
ID and CCLE names. Contains: Genes 0, Cell Lines 1677, Primary Diseases 38, 
Lineages 33. Columns are "DepMap_ID", "CCLE_Name", "Aliases", "COSMIC_ID", 
"Sanger.ID", "Primary.Disease", "Subtype.Disease", "Gender" and "Source"  

```{r, eval=FALSE, echo=T}
data("depmap_19Q1_cell_lines", package = "depmap")
depmap_19Q1_cell_lines
print("depmap_19Q1_cell_lines data column names")
names(depmap_19Q1_cell_lines)
View(depmap_19Q1_cell_lines)
```

### "depmap_19Q1_mutation_calls" 
Merged mutation calls (coding region, germline filtered). Contains: Genes 18755, 
Cell Lines 1601, Primary Diseases 37, Lineages 33. Columns are: "X1", 
"Hugo_Symbol", "Entrez_Gene_Id", "NCBI_Build", "Chromosome", "Start_position",        
"End_position", "Strand", "Variant_Classification", "Variant_Type", 
"Reference_Allele", "Tumor_Seq_Allele1", "dbSNP_RS", "dbSNP_Val_Status", 
"Genome_Change", "Annotation_Transcript", "Tumor_Sample_Barcode", "cDNA_Change",
"Codon_Change", "Protein_Change", "isDeleterious", "isTCGAhotspot", "TCGAhsCnt",
"isCOSMIChotspot", "COSMIChsCnt", "ExAC_AF", "VA_WES_AC", "CGA_WES_AC", 
"SangerWES_AC", "SangerRecalibWES_AC", "RNAseq_AC", "HC_AC", "RD_AC", "WGS_AC", 
"Variant_annotation", "DepMap_ID"  

```{r, eval=FALSE, echo=T}
data("depmap_19Q1_mutation_calls", package = "depmap")
depmap_19Q1_mutation_calls
print("depmap_19Q1_mutation_calls data column names")
names(depmap_19Q1_mutation_calls)
View(depmap_19Q1_mutation_calls)
```

### "gene_effect_corrected"
Batch corrected CERES inferred gene effect matrix derived from CRISPR knock-out
data. Contains: Genes 17634, Cell Lines 558, Primary Diseases 26, Lineages 28. 
First column is the DepMap ID# of the cell lineage. The remaining columns 
consist of the numerical dependency score values of are select genes. 

```{r, eval=FALSE, echo=T}
data("gene_effect_corrected", package = "depmap")
gene_effect_corrected
print("gene_effect_corrected data subset column names 1-10")
names(gene_effect_corrected[c(1:10)])
View(gene_effect_corrected[c(1:10)])
```

### "public_19Q1_gene_cn" 
DepMap WES copy number data. Contains: Genes 23299, Cell Lines 1604, Primary 
Diseases 38, Lineages 33. First column is the DepMap ID# of the cell lineage. 
Columns with numerical dependency score values of are select genes. 

```{r, eval=FALSE, echo=T}
data("public_19Q1_gene_cn", package = "depmap")
public_19Q1_gene_cn
print("public_19Q1_gene_cn data subset column names 1-10")
names(public_19Q1_gene_cn[c(1:10)])
View(public_19Q1_gene_cn[c(1:10)])
```

## Downloading [Broad Institute](https://depmap.org/portal/download/) data directly

If desired, the [depmap](https://github.com/tfkillian/depmap/) datasets can also 
be downloaded in *.csv* format directly from the [Broad Institute](https://depmap.org/portal/download/) 
website. However, be aware that these files are 1.5GB in total (non-compressed) 
and take a moderate amount of time to download.

```{r, eval=FALSE, echo=T}
library("RCurl", warn.conflicts = FALSE)

# r read csv from url allows direct download csv file from website

# download depmap_19Q1_mutation_calls.csv
depmap_19Q1_mutation_calls <- read.csv("https://depmap.org/portal/download/api/download/external?file_name=ccle%2Fdepmap-mutation-calls-9a1a.7%2Fdepmap_19Q1_mutation_calls.csv")
depmap_19Q1_mutation_calls

# download public_19Q1_gene_cn.csv
public_19Q1_gene_cn  <- read.csv("https://depmap.org/portal/download/api/download/external?file_name=ccle%2Fdepmap-wes-cn-data-97cc.14%2Fpublic_19Q1_gene_cn.csv")
public_19Q1_gene_cn

# download DepMap-2019q1-celllines.csv
depmap_19Q1_cell_lines  <- read.csv("https://depmap.org/portal/download/api/download/external?file_name=processed_portal_downloads%2Fdepmap-public-cell-line-metadata-183e.3%2FDepMap-2019q1-celllines.csv")
depmap_19Q1_cell_lines

# download gene_effect_corrected.csv
gene_effect_corrected <- read.csv("https://ndownloader.figshare.com/files/14221385")
gene_effect_corrected

# download CCLE_depMap_19Q1_TPM.csv
CCLE_depMap_19Q1_TPM  <- read.csv("https://depmap.org/portal/download/api/download/external?file_name=ccle%2Fdepmap-rnaseq-expression-data-ccd0.12%2FCCLE_depMap_19Q1_TPM.csv")
CCLE_depMap_19Q1_TPM

# download CCLE_RPPA_20180123.csv
CCLE_RPPA_20180123 <- read.csv("https://depmap.org/portal/download/api/download/external?file_name=ccle%2FCCLE_RPPA_20180123.csv")
CCLE_RPPA_20180123

# download D2_combined_genetic_dependency_scores.csv
D2_combined_genetic_dependency_scores <- read.csv("https://ndownloader.figshare.com/files/13515395")
D2_combined_genetic_dependency_scores
```

## Visualisations

Below are a few plots to help illustrate features of the [depmap](https://github.com/tfkillian/depmap/) 
data. The data loaded in the [depmap](https://github.com/tfkillian/depmap/) 
package is taken from a diverse array of cancer lineages, intended to represent 
the full spectrum of fatal cancers. 

```{r, echo=FALSE, fig.width=10}
library(ggplot2, warn.conflicts = FALSE) 
library(viridis, warn.conflicts = FALSE)

#select data
cancer_types <-unique(depmap_19Q1_cell_lines$Primary.Disease)
num_cancer <-(as.integer(cancer_types))

#plot pie
pie(x=num_cancer, label = "", init.angle = 90, col = plasma(length(cancer_types)), 
    main = "Proportion of 501 Cancer Lineages Represented in DepMap Dependency Screen")
legend("left", legend=cancer_types, fill=plasma(length(cancer_types)))

```

Much of the data that can be found within the [depmap](https://github.com/tfkillian/depmap/) 
package relates to the dependency score. This score represents the observed 
shRNA log fold change (LFC) depletion values in each cell line (CL) as a 
combination of gene knockdown and off-target seed effects. Or in other words, 
the dependency score expresses how vital a particular gene is in terms of how 
lethal the knockout of that gene is on the target cell line. A negative 
dependency score is indicative of high lethality, while a positive score 
indicates improved CL vitality. A dependency score of zero indicates no effect.   

The function below fetches the dependency scores corresponding to a specific 
gene and cell lineage. For example for an example gene we will use `BRCA1` a
well known human tumor suppressor gene and compare with the `184A1_BREAST` 
breast cancer lineage.

```{r, eval=TRUE, echo=T}

# gives a logical vector isolating gene in first column
BRCA1_row <- D2_combined_genetic_dependency_scores$X1=="BRCA1 (672)"

#select row and column
ds_brca1_breast <- D2_combined_genetic_dependency_scores[BRCA1_row, "184A1_BREAST"]

#print dependency score
num_brca1_breast <- as.numeric(ds_brca1_breast[1,1])
BRCA1_val <- paste("Dependency score for is BRCA1 on 184A1_BREAST CL is: ", num_brca1_breast)
print(BRCA1_val)
# "Dependency score for is BRCA1 on 184A1_BREAST CL is:  0.0144140222477"
```

As can be seen here dependency score of the knockout of BRCA1 is slightly 
positive, which implies that that removal of this gene is not lethal for this 
breast cancer lineage, which is what we would predict. However, we would like 
to look at the highest dependency scores for this lineage. 

If we want to compare the average dependency score for all genes for the 
`184A1_BREAST` cell line:

```{r, eval=TRUE, echo=T}
#remove NA values which prevent calculation of mean
no_na_breast <-na.omit(D2_combined_genetic_dependency_scores$`184A1_BREAST`)

#avg gene dependency
breast_avg <- mean(no_na_breast)
breast_val <- paste("The average dependency score for all genes for the 184A1_BREAST CL is: ", breast_avg)
print(breast_val)
# "The average dependency score for all genes for 184A1_BREAST CL is:  -0.03864401760922"
```

This implies that the knockout of an average gene for this cell line is mildly 
deleterious, but not fatal.

The most interesting genes are the ones with the lowest dependency score, which
implies that they are the most vital to the viability of that particular lineage
and the removal of that gene is likely fatal. Such genes are consequently 
desirable as potential drug targets.

Given some gene, can we tell if it's vital or not ? How can we evaulate? Below 
we will also plot for all genes, vs avg value and then isolate highest effect.

```{r, eval=FALSE, echo=T}
#sort dependencies scores of particular cell line my lowest (most important)
h <-D2_combined_genetic_dependency_scores[order(D2_combined_genetic_dependency_scores$`184A1_BREAST`, decreasing=F)[1:10],]
dep_values_breast <- rbind(h$X1, h$`184A1_BREAST`)
print("Dependency scores for top 10 dependent genes for 184A1_BREAST CL")
dep_values_breast

#[1] "Dependency scores for top 10 dependent genes for 184A1_BREAST CL"
#     [,1]             [,2]              [,3]             [,4]          [,5]             [,6]              [,7]             
#[1,] "SF3B1 (23451)"  "MMS22L (253714)" "VARS (7407)"    "RPL6 (6128)" "EEF1A1 (1915)"  "COLEC10 (10584)" "CASP8AP2 (9994)"
#[2,] "-2.29742951572" "-2.0783201785"   "-2.03016402347" "-1.96385881" "-1.93895282018" "-1.91599928102"  "-1.84969637225" 
#     [,8]             [,9]             [,10]           
#[1,] "SF3B2 (10992)"  "EIF3B (8662)"   "TRRAP (8295)"  
#[2,] "-1.82570314454" "-1.81781811554" "-1.73550020792"

```

The gene with the highest dependency score, `SF3B1` Entrez Gene ID: (23451) 
encodes subunit 1 of the splicing factor 3b protein complex.

If we wish to capture the entire dependency map of the `184A1_BREAST` cell line:

```{r, eval=FALSE, echo=T}
library(ggplot2, warn.conflicts = FALSE) 
library(viridis, warn.conflicts = FALSE)
library(tidyverse, warn.conflicts = FALSE)

#omit genes with NA dependency
no_na_breast <-na.omit(D2_combined_genetic_dependency_scores$`184A1_BREAST`)

#coerce into data frame
no_na_breast <- as.data.frame(no_na_breast)
no_na_breast <- tibble::rowid_to_column(no_na_breast, "ID")
Breast_184A1 <-no_na_breast

#add col names
colnames(Breast_184A1, do.NULL = FALSE)
colnames(Breast_184A1) <- c("Gene","Dependency")

#avg gene dependency
breast_avg2 <- mean(Breast_184A1$Dependency)

#plot gene dependency levels for 184A1_BREAST cell line
ggplot(Breast_184A1) + aes(Gene, Dependency, color = Breast_184A1$Dependency) + 
    geom_point() +  scale_colour_viridis() + 
    geom_hline(yintercept=breast_avg2, linetype="dashed", color = "black")

```

It might be interesting to locate the highest scoring dependencies among all 
cell lines.

```{r, eval=FALSE, echo=T}

# function that finds highest dependency scores for each gene and cell line

#find row location of minimum
#depmin <-which.min(D2_combined_genetic_dependency_scores$`127399_SOFT_TISSUE`)
#high_dep_gene <-D2_combined_genetic_dependency_scores$X1[depmin]
#high_dep_gene

#last_col <- D2_combined_genetic_dependency_scores[,ncol(D2_combined_genetic_dependency_scores)]
# number is 713 columns

low_dep_scores <- apply(D2_combined_genetic_dependency_scores[c(2:713)], 2, 
                        function(x) {min(x, na.rm = TRUE)})

low_dep_gene <- apply(D2_combined_genetic_dependency_scores[c(2:713)], 2, 
                function(x) {depmin <-which.min(x)
                D2_combined_genetic_dependency_scores$X1[depmin]})

# coerce into data frame
low_dep_scores <- as.data.frame(low_dep_scores)
low_dep_gene <- as.data.frame(low_dep_gene)

# turn row names into a column
r1 <-rownames(low_dep_scores)
low_dep_scores$Cell_Line <- r1
r2 <-rownames(low_dep_gene)
low_dep_gene$Cell_Line <- r2

# tidy up df, merge on Cell_Line
df3 = merge(low_dep_gene, low_dep_scores, by.x=c("Cell_Line"), 
            by.y=c("Cell_Line"))
colnames(df3, do.NULL = FALSE)
colnames(df3) <- c("Cell_Line", "Gene", "Dependency_Score")

# sort by 10 highest dependencies 
top10 <-df3[order(df3$Dependency_Score, decreasing=F)[1:10],]
top10
```

The gene with the highest dependency score is UBC (7316). This gene represents a 
ubiquitin gene, ubiquitin C. The encoded protein is a polyubiquitin precursor.  
Ubiquitination has been associated with protein degradation, DNA repair, cell 
cycle regulation, kinase modification, endocytosis, and regulation of other cell 
signaling pathways.

If we look for the highest dependency for the SW1088_CENTRAL_NERVOUS_SYSTEM cell
line, we see indeed that UBC (7316) is the most dependent gene for vitality. 

```{r, eval=FALSE, echo=T}
#sort dependencies scores of particular cell line my lowest (most important)
ubc <-D2_combined_genetic_dependency_scores[order(D2_combined_genetic_dependency_scores$SW1088_CENTRAL_NERVOUS_SYSTEM, decreasing=F)[1:10],]
dep_values_cns <- rbind(ubc$X1, ubc$SW1088_CENTRAL_NERVOUS_SYSTEM)
print("Dependency scores for top 10 dependent genes for SW1088_CENTRAL_NERVOUS_SYSTEM")
dep_values_cns

# [1] "Dependency scores for top 10 dependent genes for SW1088_CENTRAL_NERVOUS_SYSTEM"
#     [,1]             [,2]             [,3]             [,4]             [,5]             [,6]            
# [1,] "UBC (7316)"     "PSMA3 (5684)"   "PSMA4 (5685)"   "UBA52 (7311)"   "COPZ1 (22818)"  "SRSF1 (6426)"  
# [2,] "-5.93237872763" "-3.17473770863" "-2.95412602267" "-2.86619955206" "-2.68313273074" "-2.63257840872"
#     [,7]             [,8]            [,9]             [,10]                           
# [1,] "NACA (4666)"    "MED28 (80306)" "UBB (7314)"     "U2AF1L5&U2AF1 (102724594&7307)"
# [2,] "-2.52727165132" "-2.4743706891" "-2.31711569988" "-2.30982469365"        

```

Below is a plot of the lowest dependency scores for each gene and CL for the 
`D2_combined_genetic_dependency_scores` dataset. 
 
```{r, eval=FALSE, echo=T}
# apply function to get mean dependency from all columns
D2_avg <- apply(D2_combined_genetic_dependency_scores[c(2:713)], 2, 
                        function(x) {mean(x, na.rm = TRUE)})
D2_avg2 <- mean(D2_avg)
# -0.06833473

# apply function to get mean HIGHEST dependency from all columns
df3_avg <- mean(df3$Dependency_Score)
# -2.198807

#coerce into data frame
df3 <- as.data.frame(df3)
#df3 <- tibble::rowid_to_column(df3, "ID")
#View(df3)

#plot gene dependency levels for all cell lines
ggplot(df3) + aes(ID, Dependency_Score, color = df3$Dependency) + 
    geom_point() +  scale_colour_viridis() + 
    geom_hline(yintercept=D2_avg2, linetype="dashed", color = "black") + 
    geom_hline(yintercept=df3_avg, linetype="dashed", color = "red") +
    labs(y = "Largest Dependency Score", x = "Genes") + 
    ggtitle("Plot of the greatest dependency scores") +
    scale_linetype_manual(name = "Avg dep scores",values = c(2,2))

# I will add a legend to the lines when I have time

```

It might be insightful to compare the dependency scores of datasets derived from
RNAi knockout e.g. `D2_combined_genetic_dependency_scores` and those utilizing 
CRISPR knockouts e.g. `gene_effect_corrected` . 

```{r, eval=FALSE, echo=T}

gene_effect_corrected

D2_combined_genetic_dependency_scores

```



(1) replicate one or two figures as produced by the online DepMap Data Explorer 
portal

```{r, eval=FALSE, echo=T}

gene_effect_corrected

D2_combined_genetic_dependency_scores

```

(2) compute and visualise initially for a single gene the correlation between 
shRNAi and CRISPR-based dependency scores

```{r, eval=FALSE, echo=T}

```

(3) compute and visualise for all genes, the correlation between shRNAi and 
CRISPR-based dependency scores

```{r, eval=FALSE, echo=T}

```

(4) explore the dependency scores for genes with high and low expression 

PSMC2 (Rpt1) was the highest-ranked CYCLOPS candidate gene in
our original analysis and was also significant in the validation
data set. PSMC2 is part of the 19S regulatory complex of the
26S proteasome, which is responsible for catalyzing the unfold-
ing and translocation of substrates into the 20S proteasome

```{r, eval=FALSE, echo=T}

```


